Concurrent Programming with CXL

### 2020-06-24 book/paper.tex
Now supports += etc

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def incrementer(self):
   0 Jump 32
   1 Frame incrementer(self) end=9
Up.cxl:2     count = count + 1;
   2 Push 1
   3 Load count
   4 2-ary +
   5 Store count
Up.cxl:3     done[self] = True;
   6 Push True
   7 PushVar self
   8 Store done 1
   9 Return
\end{verbatim}
\end{code}
\caption{The first part of the CVM bytecode corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
#states = 79
==== Safety violation ====
CXL Assertion failed ('assert', 'Up.cxl', 9, 5) 1
__init__/() [0,32-47]       48 dict{ .count:0, .done:dict{ 0:False, 1:False } }
incrementer/0 [1-4]          5 dict{ .count:0, .done:dict{ 0:False, 1:False } }
incrementer/1 [1-8]          9 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
incrementer/0 [5-8]          9 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
main/() [11-15,18-22,24-29] 29 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

\begin{enumerate} \setcounter{enumi}{1}
\item the \texttt{Push} instruction pushes the constant 1
onto the stack of the process.
\item The \texttt{Load} instruction pushes the value of the
\texttt{count} variable onto the stack.
\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
It pops two values from the stack (1 and the value of \texttt{count}),
adds them and pushes the result back onto the stack.
\item The \texttt{Store} instruction pops pops
a CXL value (the sum of the \texttt{count} variable and 1) and
stores it in the \texttt{count} variable.
\end{enumerate}

@@

If we look at the middle three rows, we see that:
\begin{itemize}
\item Process 0 executed instructions 1 through 4, loading the value of
\texttt{count} but stopping just before storing 1 into \texttt{count};
\item Process 1 executed instructions 1 through 8, storing 1 into
into \texttt{count} and storing \texttt{True} into \texttt{done[1]};
\item Process 0 continues execution, storing value 1 into \texttt{count}
and storing \texttt{True} into \texttt{done[0]}.
\end{itemize}

@@

\begin{figure}
\begin{center}
\includegraphics[width=4.5in]{figures/mutex-crop.pdf}
\end{center}
\caption{High-level state diagram specification of mutual exclusion with up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes in the critical section.}
\label{fig:mutex}
\end{figure}

@@

While the CVM does not have any specific built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} statement blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
atomic_inc(&count);
\end{Verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def Semaphore(cnt):
    result = cnt;
;
def P(sema):
    let blocked = True:
        while blocked:
            atomic:
                if (^sema) > 0:
                    ^sema = (^sema) - 1;
                    blocked = False;
                ;
            ;
        ;
    ;
;
def V(sema):
    atomic:
        ^sema = (^sema) + 1;
    ;
;
\end{Verbatim}
\end{code}
\caption{The \texttt{Semaphore} interface in the \texttt{synch} module.}
\label{fig:semaphore}
\end{figure}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & \texttt{False}, \texttt{True} \\
\hline
integers & \texttt{..., -2, -1, 0, 1, 2, ...} \\
\hline
Atom & \texttt{.example}, \texttt{.test1} \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
\hline
Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
\hline
Address & \texttt{\&lock, \&flags[2]} \\
\hline
Context & (generated by \texttt{stop} expression) \\
\hline
\end{tabular}
\vspace{1em}

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item atoms are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of atoms.  Addresses are lexicographically
ordered accordingly;
\item contexts (Appendix~\ref{app:context}) are ordered by hash.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{bagEmpty()} & returns an empty bag\\
\hline
\texttt{bagFromset(s)} & create a bag from set \texttt{s}\\
\hline
\texttt{bagCount(b, e)} & count how many times \texttt{e} occurs in bag \texttt{b}\\
\hline
\texttt{bagChoose(b)} & like \texttt{choose(s)}, but applied to a bag\\
\hline
\texttt{bagAdd(pb, e)} & add one copy of \texttt{e} to bag \texttt{\^{}pb}\\
\hline
\texttt{bagRemove(pb, e)} & remove one copy of \texttt{e} from bag \texttt{\^{}pb}\\
\hline
\end{tabular}
\vspace{1em}

>>>>

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def incrementer(self):
   0 Jump 34
   1 Frame incrementer(self) end=11
Up.cxl:2     count = count + 1;
   2 Load count
   3 Push 1
   4 2-ary +
   5 Store count
Up.cxl:3     done[self] = True;
   6 LoadVar self
   7 PushAddress done
   8 Address 2
   9 Push True
   10 Store
   11 Return
\end{verbatim}
\end{code}
\caption{The first part of the CVM bytecode corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
#states = 79
==== Safety violation ====
CXL Assertion failed ('assert', 'Up.cxl', 9, 5): 1
__init__/() [0,34-49]       50 dict{.count:0, .done:dict{0:False, 1:False}}
incrementer/0 [1-4]          5 dict{.count:0, .done:dict{0:False, 1:False}}
incrementer/1 [1-10]        11 dict{.count:1, .done:dict{0:False, 1:True }}
incrementer/0 [5-10]        11 dict{.count:1, .done:dict{0:True,  1:True }}
main/() [13-17,20-24,26-31] 31 dict{.count:1, .done:dict{0:True,  1:True }}
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

\begin{enumerate} \setcounter{enumi}{1}
\item The \texttt{Load} instruction pushes the value of the
\texttt{count} variable onto the stack.
\item the \texttt{Push} instruction pushes the constant 1
onto the stack of the process.
\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
It pops two values from the stack (1 and the value of \texttt{count}),
adds them and pushes the result back onto the stack.
\item The \texttt{Store} instruction pops
a CXL value (the sum of the \texttt{count} variable and 1) and
stores it in the \texttt{count} variable.
\end{enumerate}

@@

If we look at the middle three rows, we see that:
\begin{itemize}
\item Process 0 executed instructions 1 through 4, loading the value of
\texttt{count} but stopping just before storing 1 into \texttt{count};
\item Process 1 executed instructions 1 through 10, storing 1 into
into \texttt{count} and storing \texttt{True} into \texttt{done[1]};
\item Process 0 continues execution, storing value 1 into \texttt{count}
and storing \texttt{True} into \texttt{done[0]}.
\end{itemize}

@@

\begin{figure}
\begin{center}
\includegraphics[width=4in]{figures/mutex-crop.pdf}
\end{center}
\caption{High-level state diagram specification of mutual exclusion with up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes in the critical section.}
\label{fig:mutex}
\end{figure}

@@

While the CVM does not have any specific built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} statement blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr += 1;
    ;
;
count = 0;
atomic_inc(&count);
\end{Verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def Semaphore(cnt):
    result = cnt;
;
def P(sema):
    let blocked = True:
        while blocked:
            atomic:
                if (^sema) > 0:
                    ^sema -= 1;
                    blocked = False;
                ;
            ;
        ;
    ;
;
def V(sema):
    atomic:
        ^sema += 1;
    ;
;
\end{Verbatim}
\end{code}
\caption{The \texttt{Semaphore} interface in the \texttt{synch} module.}
\label{fig:semaphore}
\end{figure}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & \texttt{False}, \texttt{True} \\
\hline
integers & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
\hline
Atom & \texttt{.example}, \texttt{.test1} \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
\hline
Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
\hline
Address & \texttt{\&lock, \&flags[2], NULL} \\
\hline
Context & (generated by \texttt{stop} expression) \\
\hline
\end{tabular}
\vspace{1em}

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item atoms are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of atoms.  \texttt{NULL} is the empty list of atoms.
Addresses are lexicographically ordered accordingly;
\item contexts (Appendix~\ref{app:context}) are ordered by hash.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{bagEmpty()} & returns an empty bag\\
\hline
\texttt{bagFromSet(s)} & create a bag from set \texttt{s}\\
\hline
\texttt{bagCount(b, e)} & count how many times \texttt{e} occurs in bag \texttt{b}\\
\hline
\texttt{bagChoose(b)} & like \texttt{choose(s)}, but applied to a bag\\
\hline
\texttt{bagAdd(pb, e)} & add one copy of \texttt{e} to bag \texttt{\^{}pb}\\
\hline
\texttt{bagRemove(pb, e)} & remove one copy of \texttt{e} from bag \texttt{\^{}pb}\\
\hline
\end{tabular}
\vspace{1em}
